dispRity manual

### 2017-08-04 doc/dispRity-manual.Rmd
Fixed typos etc to 'make.metric'

The function `time.subsamples` allows to divide the ordinated matrix into different time subsamples given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
Each subsample generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.

@@

For the time-slicing method details see @GuillermeSTD.
For both methods, the function intakes the `time` argument which can be a vector of `numeric` values for:

@@

Otherwise, the `time` argument can be set as a single `numeric` value for automatically generating a given number of equidistant time-bins/slices.
Additionally, it is also possible to input a data frame containing the first and last occurrence data (FAD/LAD) for taxa that span over a longer time than the given tips/nodes age.

@@

```{r, eval=TRUE}
## Generating three time bins containing the data present every 40 Ma
time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
                time = c(120, 80, 40, 0))
```

Note that we can also generate equivalent results by just telling the function that we want three time-bins (subsamples) as follow:

@@

In this example, the taxa were split inside each time-bin according to their age.
However, the taxa here are considered as single points in time.
It is totally possible that some taxa could have had longer longevity and that it spans across multiple time bins.
In this case, it is possible to include them in more than one bin by providing a table of first and last occurrence dates (FAD/LAD).
This table should have the taxa names as row names and two columns for respectively the first and last occurrence age:

@@

When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age plus 1% of the tree length as long as at least three elements are present at that point in time.
The algorithm adds an extra 1% tree length until reaching the required minimum of three elements.
It is also possible to include nodes in each bin by using `inc.nodes = TRUE` and providing a matrix that contains the ordinated distance between tips *and* nodes.

For the time-slicing method (`method = continuous`), the idea is fairly similar.
This option, however, requires a matrix that contains the ordinated distance between taxa *and* nodes as well as an extra argument being the assumed evolutionary model (via the `model` argument).
This model argument is used when the time slice occurs along a branch of the tree rather than on a tip or a node.
The model can be one of the following:

@@

```{r, eval=TRUE}
## Generating four time slices every 40 million years under gradual evolution
time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages)

@@

@@

One important step in analysing ordinated matrices is to pseudo-replicate the data to see how robust the results of future analysis are sensitive to outliers in the dataset.
This can be achieved using the function `boot.matrix` to bootstrap or/and rarefy the data.
The default options will bootstrapping the matrix 100 times without rarefaction using the "full" bootstrap method (see below):

@@

This function also allows to rarefy the data using the `rarefaction` argument.
Rarefaction allows to limit the number of elements to be drawn at each bootstrap replication.
For example, one might be interested in looking at the effect of reducing the number of elements on the results of an analysis.

This can be achieved by using the `rarefaction` option that draws only *n-x* at each bootstrap replicate (where *x* is the number of elements not sampled).
The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e. remove *x* elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subsamples to a minimum of 3 elements).
It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.

@@

## Or with a set number of rarefaction levels
boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8,3))
```

One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
When left missing, all dimensions from the ordinated matrix are used.

```{r, eval=TRUE}
## Removing the last axis 50% last dimensions
boot.matrix(BeckLee_mat50, dimensions = 0.5)

## Keeping the 10 first dimensions
boot.matrix(BeckLee_mat50, dimensions = 10)
```

Of course, one could be interested in directly supplying the subsamples generated above directly to this function.
In fact, it can also deal with a list of matrices or with a `dispRity` object output from the `custom.subsamples` or `time.subsamples` functions.

@@

## Creating time slices subsamples
time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages)
## Bootstrapping on the time binning/slicing subsamples
boot.matrix(time_slices, bootstraps = 100)
```

@@

There are many ways for measuring disparity!
In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
For example, one can look at ellipsoid hyper-volume of the ordinated space @DonohueDim, the sum and the product of the ranges and variances @Wills1994 or the median position of the elements relative to their centroid @finlay2015morphological.
Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks or some being more generalist.

Because of this great diversity of metric, the package `dispRity` has not a unique way to measure disparity but rather proposes to facilitate users to define their own disparity metric that will best suit to their particular analysis.
In fact, the core function of the package, `dispRity`, allows the user to define any metric with the `metric` argument.
However the `metric` argument has to follow certain rules:

1.  it must be composed from one to three `function` objects;
2.  the(se) function(s) must intake as a first argument a `matrix` or a `vector`;
3.  the(se) function(s) must be of one of the three levels described below;
4.  at least one of the functions must be of level 1 or 2 (see below).

### The function levels

The metric functions levels determine the "dimensionality" level of decomposition of the input ordinated matrix.
In other words, each level designate the dimensions of the output: either three (a `matrix`); two (a `vector`); or one (a single `numeric` value) dimensions.

![Illustration of the different levels of output of a function with an input `matrix`](dispRity_fun.png)

#### A level 1 function

A level 1 function will decompose a `matrix` or a `vector` into a single value:

@@

## Example of level 1 functions
mean(dummy_matrix)
median(dummy_matrix)
```

Any summary metric such as mean or median are good examples of level 1 functions as they reduce the matrix to a single dimension (i.e. one value).

#### A level 2 function

A level 2 function will decompose a `matrix` into a `vector`.

@@

## A level 2 metric
prod.rows(dummy_matrix)
```

Several level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` or `ranges` functions that calculates the variance or the range of each dimension of the ordinated matrix respectively.

#### A level 3 function

Finally a level 3 function will transform the matrix into another matrix.
Note that the dimension of the output matrix don't need to match the the input one:

```{r}
## A level 3 metric
var(dummy_matrix)

## A level 3 metric with a forced matrix output
as.matrix(dist(dummy_matrix))
```

@@

Of course, functions can be more complex and involve multiple operations such as the `centroids` function (see `?dispRity.metric`) that calculates the euclidean distance between each elements and the a position in the ordinated space.
The `make.metric` function implemented in `dispRity` is designed to help testing and finding the level of the functions.
This function tests:

1.  if your function can deal with a `matrix` or a `vector` as an input;
2.  which is your function's level according to it's output (level 1, 2 or 3, see above);
3.  whether the function can properly be implemented in the `dispRity` function (the function is fed into a `lapply` loop).

For example, let's see if the functions described above are the right levels:

```{r}
## Which level is the mean function? And is able to be used in the dispRity?
make.metric(mean)

@@

A silent version of the function is also available to be included in your code.
This can be done using the option `silent = TRUE` and will simply output the level of the metric.

```{r}
## Testing whether mean is level 1
if(make.metric(mean, silent = TRUE) != "level1") {
    message("The metric is not level 1.")
}
## Testing whether var is level 1
if(make.metric(var, silent = TRUE) != "level1") {
    message("The metric is not level 1.")
}
```

@@

```{r}
## Measuring disparity as the standard deviation of all the value of the
## ordinated matrix (level 1 function).
summary(dispRity(BeckLee_mat50, metric = sd))

## Measuring disparity as the standard deviation of the variance of each axis of
## the ordinated matrix (level 1 and level 2 functions).
summary(dispRity(BeckLee_mat50, metric = c(sd, variances)))

## Measuring disparity as the standard deviation of the variance of each axis of
## the variance covariance matrix (level 1, level 2 and level 3 functions).
summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
```

Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function levels (using `make.metric`) and apply them to the data in decreasing order (level 3 > level 2 > level 1).

@@

This is possible in the `dispRity` package by calculating disparity as a level 2 metric only!
Let's have a look using our previous example of bootstrapped time slices but by measuring the distances between each taxa and their centroid as disparity

@@

The resulting disparity object being a level 2, it can easily also be transformed into a level 1 one by for example measuring the median distance of all these distributions:

>>>>

The function `time.subsamples` allows users to divide the ordinated matrix into different time subsamples or slices given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
Each subsample generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.

@@

For the time-slicing method details see @GuillermeSTD. 
<!-- NC: Or potentially this paper we are writing for the PalAss? -->
For both methods, the function takes the `time` argument which can be a vector of `numeric` values for:

@@

Otherwise, the `time` argument can be set as a single `numeric` value for automatically generating a given number of equidistant time-bins/slices.
Additionally, it is also possible to input a dataframe containing the first and last occurrence data (FAD/LAD) for taxa that span over a longer time than the given tips/nodes age, so taxa can appear in more than one time bin/slice.

@@

```{r, eval=TRUE}
## Generating three time bins containing the taxa present every 40 Ma
time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
                time = c(120, 80, 40, 0))
```

Note that we can also generate equivalent results by just telling the function that we want three time-bins as follow:

@@

In this example, the taxa were split inside each time-bin according to their age.
However, the taxa here are considered as single points in time.
It is totally possible that some taxa could have had longer longevity and that they exist in multiple time bins.
In this case, it is possible to include them in more than one bin by providing a table of first and last occurrence dates (FAD/LAD).
This table should have the taxa names as row names and two columns for respectively the first and last occurrence age:

@@

When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as the root age plus 1\% of the tree length, as long as at least three elements/taxa are present at that point in time.
The algorithm adds an extra 1\% tree length until reaching the required minimum of three elements.
It is also possible to include nodes in each bin by using `inc.nodes = TRUE` and providing a matrix that contains the ordinated distance among tips *and* nodes.

For the time-slicing method (`method = continuous`), the idea is fairly similar.
This option, however, requires a matrix that contains the ordinated distance among taxa *and* nodes and an extra argument describing the assumed evolutionary model (via the `model` argument).
This model argument is used when the time slice occurs along a branch of the tree rather than on a tip or a node, meaning that a decision must be made about what the value for the branch should be.
The model can be one of the following:

@@

<!--NC: Bit puzzled by the offspring/descendent terminology - I'd assume these were the same thing. By descendant do you mean ancestor? I'm confused  -->

```{r, eval=TRUE}
## Generating four time slices every 40 million years under a model of gradual evolution
time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages)

@@

<!-- NC: Their idea? Huh? Whats that mean? -->

@@

One important step in analysing ordinated matrices is to pseudo-replicate the data to see how robust the results are, and how sensitive they are to outliers in the dataset.
This can be achieved using the function `boot.matrix` to bootstrap and/or rarefy the data.
The default options will bootstrap the matrix 100 times without rarefaction using the "full" bootstrap method (see below):

@@

This function also allows users to rarefy the data using the `rarefaction` argument.
Rarefaction allows users to limit the number of elements to be drawn at each bootstrap replication.
This is useful if, for example, one is interested in looking at the effect of reducing the number of elements on the results of an analysis.

This can be achieved by using the `rarefaction` option that draws only *n-x* at each bootstrap replicate (where *x* is the number of elements not sampled).
The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e. remove *x* elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subsample to a minimum of three elements).
It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.

@@

## Or with a set number of rarefaction levels
boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8, 3))
```

One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
When missing, all dimensions from the ordinated matrix are used.

<!-- NC: This is confusing to me. If it's a number > 0, it keeps that number of dimensions, and if < 0 it removes the last dimensions by that %? This is very odd. Please explain -->

```{r, eval=TRUE}
## Removing the last 50% of the dimensions
boot.matrix(BeckLee_mat50, dimensions = 0.5)

## Keeping the first 10 dimensions
boot.matrix(BeckLee_mat50, dimensions = 10)
```

Of course, one could directly supply the subsamples generated above (using <!--  NC: which function ? How is this line different from the statement in the line below? -->) to this function.
It can also deal with a list of matrices or with a `dispRity` object output from the `custom.subsamples` or `time.subsamples` functions.

@@

## Creating time slice subsamples
time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
                               method = "continuous", model = "gradual", 
                               time = c(120, 80, 40, 0),
                               FADLAD = BeckLee_ages)
## Bootstrapping the time slice subsamples
boot.matrix(time_slices, bootstraps = 100)
```

@@

There are many ways of measuring disparity!
In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
For example, one can look at ellipsoid hyper-volume of the ordinated space @DonohueDim, the sum and the product of the ranges and variances @Wills1994 or the median position of the elements relative to their centroid @finlay2015morphological. 
<!-- NC: probably should cite Matt Wills' paper for the centroid thing too as he invented it apparently -->
Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.

Because of this great diversity of metrics, the package `dispRity` does not have one way to measure disparity but rather proposes to facilitate users in defining their own disparity metric that will best suit their particular analysis.
In fact, the core function of the package, `dispRity`, allows the user to define any metric with the `metric` argument.
However the `metric` argument has to follow certain rules:

1.  It must be composed from one to three `function` objects;
2.  The function(s) must take as a first argument a `matrix` or a `vector`;
3.  The function(s) must be of one of the three dimension-levels described below;
4.  At least one of the functions must be of dimension-level 1 or 2 (see below).

<!-- NC: We discussed this previously, but I don't like the term levels here. It's confusing because it means something specific wrt factors. Can we change it to dimension-level? That's more descriptive and clearer in my opinion. I have changed it below, see what you think -->

### The function dimension-levels

The metric function dimension-levels determine the "dimensionality of decomposition" of the input matrix.
In other words, each dimension-level designates the dimensions of the output, i.e. either three (a `matrix`); two (a `vector`); or one (a single `numeric` value) dimension.

![Illustration of the different dimension-levels of functions with an input `matrix`](dispRity_fun.png)

#### Dimension-level 1 functions

A dimension-level 1 function will decompose a `matrix` or a `vector` into a single value:

@@

## Example of dimension-level 1 functions
mean(dummy_matrix)
median(dummy_matrix)
```

Any summary metric such as mean or median are good examples of dimension-level 1 functions as they reduce the matrix to a single dimension (i.e. one value).

#### Dimension-level 2 functions

A dimension-level 2 function will decompose a `matrix` into a `vector`.

@@

## A dimension-level 2 metric
prod.rows(dummy_matrix)
```

Several dimension-level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` or `ranges` functions that calculate the variance or the range of each dimension of the ordinated matrix respectively.

#### Dimension-level 3 functions

Finally a dimension-level 3 function will transform the matrix into another matrix.
Note that the dimension of the output matrix doesn't need to match the the input matrix:

```{r}
## A dimension-level 3 metric
var(dummy_matrix)

## A dimension-level 3 metric with a forced matrix output
as.matrix(dist(dummy_matrix))
```

@@

Of course, functions can be more complex and involve multiple operations such as the `centroids` function (see `?dispRity.metric`) that calculates the Euclidean distance between each elements and the a position in the ordinated space.
The `make.metric` function implemented in `dispRity` is designed to help testing and finding the dimension-level of the functions.
This function tests:

1.  if your function can deal with a `matrix` or a `vector` as an input;
2.  which is your function's dimension-level according to it's output (dimension-level 1, 2 or 3, see above);
3.  whether the function can properly be implemented in the `dispRity` function (the function is fed into a `lapply` loop).

For example, let's see if the functions described above are the right dimension-levels:

```{r}
## Which dimension-level is the mean function? And is able to be used in the dispRity?
make.metric(mean)

@@

A silent version of the function is also available to be included in your code.
This can be done using the option `silent = TRUE` and will simply output the dimension-level of the metric.

```{r}
## Testing whether mean is dimension-level 1
if(make.metric(mean, silent = TRUE) != "dimension-level1") {
    message("The metric is not dimension-level 1.")
}
## Testing whether var is dimension-level 1
if(make.metric(var, silent = TRUE) != "dimension-level1") {
    message("The metric is not dimension-level 1.")
}
```

@@

```{r}
## Measuring disparity as the standard deviation of all the value of the
## ordinated matrix (dimension-level 1 function).
summary(dispRity(BeckLee_mat50, metric = sd))

## Measuring disparity as the standard deviation of the variance of each axis of
## the ordinated matrix (dimension-level 1 and 2 functions).
summary(dispRity(BeckLee_mat50, metric = c(sd, variances)))

## Measuring disparity as the standard deviation of the variance of each axis of
## the variance covariance matrix (dimension-level 1, 2 and 3 functions).
summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
```

Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function dimension-levels (using `make.metric`) and apply them to the data in decreasing order (dimension-level 3 > dimension-level 2 > dimension-level 1).

@@

This is possible in the `dispRity` package by calculating disparity as a dimension-level 2 metric only!
Let's have a look using our previous example of bootstrapped time slices but by measuring the distances between each taxa and their centroid as disparity

@@

The resulting disparity object being a dimension-level 2, it can easily also be transformed into a dimension-level 1 one by for example measuring the median distance of all these distributions:
